Mechanistic Modeling and Analysis of the Mammalian Unfolded Protein Response

### 2016-07-18 scientific-reports-title/Paper_v2.tex
Small change to the abstact
syntax and grammar edits

\begin{abstract}
Cells monitor protein folding by an inbuilt quality-control system in which incorrectly or misfolded folded proteins are tagged for degradation or sent back through a refolding cycle.
However, continued accumulation of incorrectly folded proteins triggers the Unfolded Protein Response (UPR), which attempts to re-establish folding homeostasis or commits the cell to apoptosis.
In this study, we developed a family of mechanistic models of the mammalian UPR system.
An ensemble of models parameters was estimated by minimizing the difference between simulations and experimental measurements using multiobjective optimization.
The ensemble of model parameters was validated using cross-validation.
Analysis of the model ensemble suggested the three branches of UPR fired simultaneously.
However, the importance of each brach was ranked ordered in time; PERK and IRE1 were more important early, while ATF6 was important later in the response.
The activity of all three branches as coordinated by the molecular chaperone BiP.
Model analysis suggested that BiP feedback was critical to the overall robustness of the system.
Removal of any one branch of BiP feedback, destabilized the other branches.
On the other hand, removal of all nodes of BiP feedback increased the overall robustness of the system.
Thus, while BiP feedback is crucial to allowing the cell to adapt to small perturbations, it also makes the system fragile and susceptible to manipulations.

@@

\section*{Introduction}
Protein folding is strategically important to cellular function in all organisms.
In eukaryotes, secreted, membrane-bound and organelle-targeted proteins are typically processed and folded in the endoplasmic reticulum (ER) \cite{naidoo2009er, ron2002translational, kaufman2002unfolded}.
Intracellular perturbations caused by a variety of stressors disturb the specialized environment of the ER leading to the accumulation of misfolded or unfolded proteins \cite{ellgaard2003qce,Fonseca:2009fk}.
Shifts in folding capacity have been associated with diseases such as cancer, diabetes and cardiovascular disorders \cite{ellgaard2003qce}. Physiological processes such as aging can also influence protein folding.
Normally, cells ensure proper protein folding using a combination of molecular chaperones, foldases and lectins \cite{naidoo2009er}.
However, when proper folding can not be restored, unfolded or misfolded proteins are targeted to ER Associated Degradation (ERAD) pathways for processing \cite{kaufman2002unfolded}.
If unfolded or misfolded proteins continue to accumulate, eukaryotes induce the unfolded protein response (UPR).
In mammalian cells, UPR is a complex signaling program mediated by three ER transmembrane receptors: activating transcription factor 6 (ATF6), inositol requiring kinase 1 (IRE1) and double-stranded RNA-activated protein kinase (PKR)-like endoplasmic reticulum kinase (PERK).
UPR performs three functions, adaptation, alarm and apoptosis. During adaptation, the UPR tries to reestablish folding homeostasis by inducing the expression of chaperones that enhance protein folding.
Simultaneously, translation is globally attenuated to reduce the ER folding load while the degradation of unfolded proteins is increased. If these steps fail, the UPR induces a cellular alarm and apoptosis program.
The alarm phase involves several signal transduction events, ultimately leading to the removal of the translational block and the down-regulation of the expression and activity of pro-survival factors such as the B-cell lymphoma 2 (Bcl2) protein.
After the alarm phase, cells can undergo apoptosis, although ER stress can also initiate autophagy \cite{ogata2006aac, yorimitsu2006ers, bernales2006ace, kamimoto2006iic, hoyerhansen2007cmc, kouroku2006esp, fujita2007ter}.
Thus, ER folding homeostasis strongly influences mammalian physiology \cite{Fonseca:2009fk}.

ER stress and UPR plays an important role in a spectrum of diseases and technological applications \cite{Chakrabarti:2011aa}.
In the context of diseases such as diabetes, pancreatic $\beta$-cells depend on efficient UPR signaling to meet the demands for constantly varying levels of insulin synthesis.
Type I diabetes is marked by excessive loss of pancreatic $\beta$-cells, while type II diabetes is marked by pancreatic $\beta$-cell dysfunction.
The large biosynthetic load placed on the ER by insulin production in response to blood glucose levels, can overwhelm the folding capacity of the ER.
This leads to PERK activation and the reduction of protein synthesis.
In PERK $\textit{-/-}$ cells, protein synthesis is unresponsive to the stress, leading to the accumulation of unfolded proteins (e.g. proinsulin) and ultimately cell death.
PERK deficient mice are more prone to diabetes and progressive hyperglycemia \cite{harding2001dma}.
In type II diabetes, ER stress leads to JNK-mediated phosphorylation of insulin receptor substrate 1 (IRS1) at S307, inhibiting insulin action \cite{ozcan2004ers}.
Nitric oxide (NO), is also a key player in $\beta$-cell death in type-I diabetes and vascular complications in type-II diabetes.
NO depletes ER Ca$^{2+}$ leading to ER stress and ultimately apoptosis.
Pancreatic $\beta$-cells have shown that NO-induced apoptosis is CHOP dependent \cite{oyadomari2001noi}.
Thus, ER stress is a critical feature of both type-I and type-II diabetes at the molecular, cellular and organismal level.
ER stress and UPR function can also be important in cancers.
The ER not only acts as the center for maturation of proteins, but also as a critical node for oxygen sensing and signaling.
In rapidly growing tumors, cells face stressors like hypoxia and nutrient deprivation both of which can lead to ER stress and ultimately UPR \cite{Koumenis:2006zr}.
Interestingly, the connection between hypoxia and UPR is through hypoxia inducible factor (HIF) independent pathways.
Hypoxia drives PERK activation and the transient phosphorylation of eIF2$\alpha$ \cite{koumenis2002regulation, ma2003delineation, koritzinsky2006gene}
on a time scale of minutes for anoxia, and much slower for hypoxic conditions \cite{koritzinsky2006gene}.
Because of their ability to transduce pro-apoptotic signals, both PERK and IRE1 have been explored as potential anti-cancer targets.
Versipelostatin, a repressor of BiP expression, has been shown to produce anti-tumor activity in MKN-74 xenograft mouse models \cite{park2004etc}.
Enhanced apoptosis has also been observed in BiP-deficient fibrosarcoma cells, and XBP1- and PERK-deficient mouse fibroblasts \cite{jamora1996itp, romero2004xbp1, bi2005esr}.
In the context of biotechnology, mammalian hosts, such as Chinese Hamster Ovary (CHO) cells, have been used for therapeutic protein production since the mid-1980s
\cite{Wurm:2004aa,Mohan:2008aa}. Several current biologics are secreted, for example, the monoclonal antibodies (MAb) interferon-$\gamma$ (IFN$\gamma$)
or erythropoeitin (EPO) \cite{Ku:2008aa}. Thus, ER processing and the unfolded protein response are critical to the production of these and many other therapeutic proteins.
However, despite its importance to both disease and biotechnology, there has been a paucity of mathematical models of UPR in mammalian systems.

@@

In this study, we developed a population of mathematical models describing the adaptation, alarm, and apoptosis phases of the mammalian unfolded protein response.
The biological connectivity of the UPR model was assembled following an extensive literature review.
The dynamics of UPR were modeled using mass action kinetics within the framework of ordinary differential equations. 
An ensemble of models parameters was estimated by minimizing the difference between simulations and experimental measurements using multiobjective optimization.
The ensemble of model parameters was validated using cross-validation.
Analysis of the model ensemble suggested the three branches of UPR fired simultaneously.
However, the importance of each brach as ranked ordered in time; PERK and IRE1 were more important early, while ATF6 was important later in the response.
The activity of all three branches as coordinated by the molecular chaperone BiP.
Model analysis suggested that BiP feedback was critical to the overall robustness of the system.
Removal of any one branch of BiP feedback, destabilized the other branches.
On the other hand, removal of all nodes of BiP feedback increased the overall robustness of the system.
Thus, while BiP feedback is crucial to allowing the cell to adapt to small perturbations, it also makes the system fragile and susceptible to manipulations.
The UPR model code and parameter ensemble is available under an MIT software license and can be downloaded from http://www.varnerlab.org.

@@

\subsection*{Formulation of the UPR network architecture}
The UPR network described the ER folding cycle, ER-associated degradation (ERAD), ER-stress transducer (PERK, IRE1 and ATF6) signaling cascades and stress-induced caspase activation (Fig. \ref{fg:fig1}).
The network consisted of 636 protein or mRNA species interconnected by 1090 interactions (Fig. \ref{fg:fig1} Inset).
Connectivity was formulated from a comprehensive review of the primary literature  \cite{naidoo2009er, ron2002translational, kaufman2002unfolded,ellgaard2003qce,Fonseca:2009fk,Schroder:2005vn,gotoh:hdc,mccullough2001gsc, belmont2008cga,Hetz:2009qc,Urano:2000uq,szegezdi2006mediators}, and from on-line databases; String-8 \cite{Jensen:2009fk}, NetworKIN \cite{Linding:2007fk} and TRANSFAC.
Model connectivity was not specific to a single cell-line; rather, it was a canonical representation of the pathways involved in monitoring and controlling the folding capacity of a generic well-mixed ER compartment.
UPR induction was modeled as the release of BiP from the ER stress transducers, PERK, IRE1$\alpha$, and ATF6 leading initially to adaptation of the folding cycle and then, subsequently, to alarm and apoptosis.
The adaption phase of UPR was marked by general translation attenuation, selective transcriptional programs for key species like bZIP transcription factor ATF4 \cite{lu2004tra}, cellular inhibitor of apoptosis (cIAP) \cite{hamanaka2008pdr}, molecular chaperones e.g., BiP \cite{harding2003isr}, and enhanced clearance of accumulated proteins via ERAD.
The alarm and apoptosis phases were mediated by the induction of CHOP \cite{ron1992cnd}, regulation of Bcl2, Bcl2-antagonist of cell death (BAD) \cite {wang1999cia} and (TNF) receptor associated factor 2 (TRAF2) \cite{lei2003jpb,putcha2003jmb, yamamoto1999bpa, szegezdi2006mediators} activation (Fig. \ref{fg:fig1}).
Model connectivity is available from the GitHub model repository, and is described further in the supplementary materials.

@@

\subsection*{The UPR model recapitulated the adaptation, alarm, and apoptotic phases of UPR}
A population of unknown model parameters was estimated from 33 dynamic and steady state data sets taken from literature (Table \ref{obj-table}).
The residual between model simulations and each of the measurement sets was simultaneously minimized using the multiobjective POETs algorithm \cite{Song:2010ij}.
A leave-eight-out cross-validation strategy was used to independently estimate the training and prediction error over the 33 data sets; we estimated four different model families,
where eight of the 33 objectives were reserved for validation and 25 were used for model training.
Starting from an initial best-fit parameter set (nominal set), more than 25,000 probable models were estimated by POETs from which we selected N = 100 models (25 from each training family) with a Pareto rank of one or less (from approximately 1200 possible choices).
The nominal, training (75 models), and prediction (25 models) errors were calculated for each objective (Table \ref{obj-table}).
Models used for prediction error calculations for a particular objective were \emph{not} trained on that objective.
The prediction likelihood was statistically significantly better for 31 of the 33 objective functions at a 95\% confidence level, compared with random parameter sets generated from the nominal set (Table \ref{obj-table}).
POETs generated model families that predicted approximately 94\% of the training data with a significantly higher likelihood than a random control.
However, the specific value of any given parameter was likely not well described; the coefficient of variation (CV) for the model parameters ranged from 0.5 - 1.6, where approximately 65\% of the parameters were constrained with a CV $\leq$ 1.0 (supplementary materials Fig. \ref{fg:Para_S1}).
The most constrained parameters involved a wide-array of functions e.g., regulation of PERK, eIF2$\alpha$, ATF4, Calcineurin, BiP, CHOP and ATF6 signaling.
However, the least constrained parameters involved JNK and apoptosis interactions.
POETs identified Pareto fronts between several objectives, e.g., O13$\times$O14, O25$\times$O29, O11$\times$O29, and O27$\times$O2, in the training data  (Fig. \ref{fg:objective_plot}).
Strong Pareto fronts suggested an inability to simultaneously model different aspects of the training data.
However, fronts could also result from experimental artifacts, e.g., variation between cell-lines, time-scale differences, or from functional relationships in the data.
Globally, adaptation and alarm phase training constraints conflicted with those involving apoptosis.
For example, objectives involving caspase-7 or caspase-9 activity conflicted with phosphorylated eIF2$\alpha$ levels.
Phosphorylation of eIF2$\alpha$ by activated PERK attenuates translation, which decreases the ER folding load.
Thus, eIF2$\alpha$ phosphorylation is a key early adaptive event in UPR. On the other hand, caspase-9 is a stress-induced death marker activated only after UPR has failed to restore folding homeostasis.
Conflicts between these early and late phase markers suggested the UPR time scale was perhaps cell-line or perturbation dependent. Next, we compared model simulations with measurements of components mediating the three phases of UPR.

%For example, XBP1 mRNA measurements (O14) conflicted with CHOP protein measurements (O13), even though these data-sets were taken from the same study and were collected in the same cell-line. XBP1 splicing increased BiP levels, which in turn reduced CHOP protein levels, hence the trade-off. Lastly, in addition to fronts, we also observed strong correlation between objectives. For example, models that performed well for the CHOP protein (O11), also performed well against Procaspase-12 (O22) measurements, even though these were not in the same cell-line or from the same study. Both CHOP and Procaspase-12 are downstream of the IRE1/TRAF2/JNK signaling cascade, so these errors were directly correlated.
%\clearpage
% \subsection*{The population of UPR models recapitulated adaptation, alarm, and apoptotic events across multiple cell-lines and timescales.}
%The counteracting effects are further substantiated in simulated knockout studies, wherein knockout of one ER stress transducer led to enhancement of the other branches of UPR (supplementary materials Fig. \ref{fg:Flux_Supp_Fig1}A-C).
%We compared the ability of the system to recuperate upon reversal of unfolded protein loads from the alarm phase (P4, Fig. \ref{fg:proof_of_concept_simulation} ) and the apoptosis phase (P2, Fig. \ref{fg:proof_of_concept_simulation} ). If we reduced the protein load after the cell was committed to apoptosis (as in P3), we found that the cell continued to function similar to the UPR state even upon protein load reduction after 25 hrs (Fig. \ref{fg:cross_plot_flux}C).

@@

\subsection*{Signal flow analysis suggested the UPR branches fired simultaneously.}
Traditionally, it has been hypothesized that there is a sequential order for firing of the ER stress transducers in UPR.
The PERK and ATF6 branches are thought to be activated before IRE1 \cite{szegezdi2006mediators} and largely promote ER adaptation to misfolding, while IRE1 transmits both survival and pro-apoptotic signals.
However, analysis of the UPR model ensemble suggested the three branches fired simultaneously, and that adaptation, alarm, or apoptosis was the result of counteracting effects of the three UPR signaling pathways (Fig. \ref{fg:proof_of_concept_simulation} and Fig. \ref{fg:Flux_Supp_Fig1}). UPR induction was controlled by manipulation of the generation rate of unfolded or misfolded protein (qP) in the ER compartment.
Upon induction, initially ($t\leq$ 1hr) the response was damped, marking the adaptation phase of UPR (Fig. \ref{fg:proof_of_concept_simulation}A).
Adaptation was followed by increased IRE1$\alpha$, PERK, and ATF6 activity at $t\sim$ 1 hr, marking the onset of the alarm phase.
The alarm phase was followed by a steady state at $t\sim$ 8-10 hrs, marking the onset of the apoptosis phase of UPR (Fig. \ref{fg:proof_of_concept_simulation}A).
Our analysis was substantiated further by looking at the fluxes at different phases of UPR induction (Fig. \ref{fg:proof_of_concept_simulation}B).
At P1, there was a marked increase in ATF4 and CHOP activity, ATF6 signaling and unfolded protein sensing and degradation by ERAD.
These are hallmarks of the adaptation-alarm phase of the UPR response (Fig. \ref{fg:cross_plot_flux}D).
The apoptosis phase at P2 ($t\geq$ 8-10 hrs) was marked by increased BiP regulation, enhanced ATF4 transcriptional activity, increased mitochondrial membrane permeability, and increased apoptotic fluxes.
On the other hand, if we reduced the load of unfolded protein in the adaptation-alarm phase (P4), the cell recuperated using its ERAD machinery and the regulation of BiP (Fig. \ref{fg:cross_plot_flux}E-F).
Thus, following initiation of UPR all three sensor branches were simultaneously activated. However, it was unclear which parameters or species controlled the responses of the three branches.
We addressed this question using sensitivity and robustness analysis.

>>>>

\begin{abstract}
Cells monitor protein folding by an inbuilt quality-control system in which incorrectly or misfolded proteins are tagged for degradation or sent back through a refolding cycle.
However, continued accumulation of incorrectly folded proteins triggers the Unfolded Protein Response (UPR), which attempts to re-establish folding homeostasis or commits the cell to apoptosis.
In this study, we developed a family of mechanistic models of the mammalian UPR system.
An ensemble of models parameters was estimated by minimizing the difference between simulations and experimental measurements using multiobjective optimization.
The ensemble of model parameters was validated using cross-validation.
Analysis of the model ensemble suggested the three branches of UPR fired simultaneously.
However, the importance of each branch was rank ordered in time; PERK and IRE1 were more important early, while ATF6 was important later in the response.
The activity of all three branches was coordinated by the molecular chaperone BiP.
Model analysis suggested that BiP feedback was critical to the overall robustness of the system.
Removal of any one branch of BiP feedback, destabilized the other branches.
On the other hand, removal of all nodes of BiP feedback increased the overall robustness of the system.
Thus, while BiP feedback is crucial to allowing the cell to adapt to small perturbations, it also makes the system fragile and susceptible to manipulation.

@@

\section*{Introduction}
Protein folding is strategically important to cellular function in all organisms.
In eukaryotes, secreted, membrane-bound, and organelle-targeted proteins are typically processed and folded in the endoplasmic reticulum (ER) \cite{naidoo2009er, ron2002translational, kaufman2002unfolded}.
Intracellular perturbations caused by a variety of stressors disturb the specialized environment of the ER leading to the accumulation of misfolded or unfolded proteins \cite{ellgaard2003qce,Fonseca:2009fk}.
Shifts in folding capacity have been associated with diseases such as cancer, diabetes, and cardiovascular disorders \cite{ellgaard2003qce}. Physiological processes such as aging can also influence protein folding.
Normally, cells ensure proper protein folding using a combination of molecular chaperones, foldases, and lectins \cite{naidoo2009er}.
However, when proper folding can not be restored, unfolded or misfolded proteins are targeted to ER Associated Degradation (ERAD) pathways for processing \cite{kaufman2002unfolded}.
If unfolded or misfolded proteins continue to accumulate, eukaryotes induce the unfolded protein response (UPR).
In mammalian cells, UPR is a complex signaling program mediated by three ER transmembrane receptors: activating transcription factor 6 (ATF6), inositol requiring kinase 1 (IRE1), and double-stranded RNA-activated protein kinase (PKR)-like endoplasmic reticulum kinase (PERK).
UPR performs three functions: adaptation, alarm, and apoptosis. During adaptation, the UPR tries to reestablish folding homeostasis by inducing the expression of chaperones that enhance protein folding.
Simultaneously, translation is globally attenuated to reduce the ER folding load while the degradation of unfolded proteins is increased. If these steps fail, the UPR induces a cellular alarm and apoptosis program.
The alarm phase involves several signal transduction events, ultimately leading to the removal of the translational block and the down-regulation of the expression and activity of pro-survival factors such as the B-cell lymphoma 2 (Bcl2) protein.
After the alarm phase, cells can undergo apoptosis, although ER stress can also initiate autophagy \cite{ogata2006aac, yorimitsu2006ers, bernales2006ace, kamimoto2006iic, hoyerhansen2007cmc, kouroku2006esp, fujita2007ter}.
Thus, ER folding homeostasis strongly influences mammalian physiology \cite{Fonseca:2009fk}.

ER stress and UPR plays an important role in a spectrum of diseases and technological applications \cite{Chakrabarti:2011aa}.
In the context of diseases such as diabetes, pancreatic $\beta$-cells depend on efficient UPR signaling to meet the demands for constantly varying levels of insulin synthesis.
Type I diabetes is marked by excessive loss of pancreatic $\beta$-cells, while type II diabetes is marked by pancreatic $\beta$-cell dysfunction.
The large biosynthetic load placed on the ER by insulin production in response to blood glucose levels, can overwhelm the folding capacity of the ER.
This leads to PERK activation and the reduction of protein synthesis.
In PERK $\textit{-/-}$ cells, protein synthesis is unresponsive to the stress, leading to the accumulation of unfolded proteins (e.g. proinsulin) and ultimately cell death.
PERK deficient mice are more prone to diabetes and progressive hyperglycemia \cite{harding2001dma}.
In type II diabetes, ER stress leads to JNK-mediated phosphorylation of insulin receptor substrate 1 (IRS1) at S307, inhibiting insulin action \cite{ozcan2004ers}.
Nitric oxide (NO), is also a key player in $\beta$-cell death in type-I diabetes and vascular complications in type-II diabetes.
NO depletes ER Ca$^{2+}$ leading to ER stress and ultimately apoptosis.
Pancreatic $\beta$-cells have shown that NO-induced apoptosis is CHOP dependent \cite{oyadomari2001noi}.
Thus, ER stress is a critical feature of both type-I and type-II diabetes at the molecular, cellular, and organismal level.
ER stress and UPR function can also be important in cancers.
The ER not only acts as the center for maturation of proteins, but also as a critical node for oxygen sensing and signaling.
In rapidly growing tumors, cells face stressors like hypoxia and nutrient deprivation both of which can lead to ER stress and ultimately UPR \cite{Koumenis:2006zr}.
Interestingly, the connection between hypoxia and UPR is through hypoxia inducible factor (HIF) independent pathways.
Hypoxia drives PERK activation and the transient phosphorylation of eIF2$\alpha$ \cite{koumenis2002regulation, ma2003delineation, koritzinsky2006gene}
on a time scale of minutes for anoxia, and much slower for hypoxic conditions \cite{koritzinsky2006gene}.
Because of their ability to transduce pro-apoptotic signals, both PERK and IRE1 have been explored as potential anti-cancer targets.
Versipelostatin, a repressor of BiP expression, has been shown to produce anti-tumor activity in MKN-74 xenograft mouse models \cite{park2004etc}.
Enhanced apoptosis has also been observed in BiP-deficient fibrosarcoma cells, and XBP1- and PERK-deficient mouse fibroblasts \cite{jamora1996itp, romero2004xbp1, bi2005esr}.
In the context of biotechnology, mammalian hosts, such as Chinese Hamster Ovary (CHO) cells, have been used for therapeutic protein production since the mid-1980s
\cite{Wurm:2004aa,Mohan:2008aa}. Several current biologics are secreted, for example, the monoclonal antibodies (MAb) interferon-$\gamma$ (IFN$\gamma$)
or erythropoeitin (EPO) \cite{Ku:2008aa}. Thus, ER processing and the UPR are critical to the production of these and many other therapeutic proteins.
However, despite its importance to both disease and biotechnology, there has been a paucity of mathematical models of UPR in mammalian systems.

@@

In this study, we developed a population of mathematical models describing the adaptation, alarm, and apoptosis phases of the mammalian unfolded protein response.
The biological connectivity of the UPR model was assembled following an extensive literature review.
The dynamics of UPR were modeled using mass action kinetics within the framework of ordinary differential equations.
An ensemble of models parameters was estimated by minimizing the difference between simulations and experimental measurements using multiobjective optimization.
The ensemble of model parameters was validated using cross-validation.
Analysis of the model ensemble suggested the three branches of UPR fired simultaneously.
However, the importance of each brach was rank ordered in time; PERK and IRE1 were more important early, while ATF6 was important later in the response.
The activity of all three branches was coordinated by the molecular chaperone BiP.
Model analysis suggested that BiP feedback was critical to the overall robustness of the system.
Removal of any one branch of BiP feedback, destabilized the other branches.
On the other hand, removal of all nodes of BiP feedback increased the overall robustness of the system.
Thus, while BiP feedback is crucial to allowing the cell to adapt to small perturbations, it also makes the system fragile and susceptible to manipulations.
The UPR model code and parameter ensemble is available under an MIT software license and can be downloaded from http://www.varnerlab.org.

@@

\subsection*{Formulation of the UPR network architecture}
The UPR network described the ER folding cycle, ER-associated degradation (ERAD), ER-stress transducer (PERK, IRE1 and ATF6) signaling cascades, and stress-induced caspase activation (Fig. \ref{fg:fig1}).
The network consisted of 636 protein or mRNA species interconnected by 1090 interactions (Fig. \ref{fg:fig1} Inset).
Connectivity was formulated from a comprehensive review of the primary literature  \cite{naidoo2009er, ron2002translational, kaufman2002unfolded,ellgaard2003qce,Fonseca:2009fk,Schroder:2005vn,gotoh:hdc,mccullough2001gsc, belmont2008cga,Hetz:2009qc,Urano:2000uq,szegezdi2006mediators}, and from on-line databases: String-8 \cite{Jensen:2009fk}, NetworKIN \cite{Linding:2007fk}, and TRANSFAC.
Model connectivity was not specific to a single cell-line; rather, it was a canonical representation of the pathways involved in monitoring and controlling the folding capacity of a generic well-mixed ER compartment.
UPR induction was modeled as the release of BiP from the ER stress transducers PERK, IRE1$\alpha$, and ATF6 leading initially to adaptation of the folding cycle and then, subsequently, to alarm and apoptosis.
The adaption phase of UPR was marked by general translation attenuation, selective transcriptional programs for key species like bZIP transcription factor ATF4 \cite{lu2004tra}, cellular inhibitor of apoptosis (cIAP) \cite{hamanaka2008pdr}, molecular chaperones e.g., BiP \cite{harding2003isr}, and enhanced clearance of accumulated proteins via ERAD.
The alarm and apoptosis phases were mediated by the induction of CHOP \cite{ron1992cnd}, regulation of Bcl2, Bcl2-antagonist of cell death (BAD) \cite {wang1999cia}, and (TNF) receptor associated factor 2 (TRAF2) \cite{lei2003jpb,putcha2003jmb, yamamoto1999bpa, szegezdi2006mediators} activation (Fig. \ref{fg:fig1}).
Model connectivity is available from the GitHub model repository and is described further in the supplementary materials.

@@

\subsection*{The UPR model recapitulated the adaptation, alarm, and apoptotic phases of UPR}
A population of unknown model parameters was estimated from 33 dynamic and steady state data sets taken from literature (Table \ref{obj-table}).
The residual between model simulations and each of the measurement sets was simultaneously minimized using the multiobjective POETs algorithm \cite{Song:2010ij}.
A leave-eight-out cross-validation strategy was used to independently estimate the training and prediction error over the 33 data sets; we estimated four different model families,
where eight of the 33 objectives were reserved for validation and 25 were used for model training.
Starting from an initial best-fit parameter set (nominal set), more than 25,000 probable models were estimated by POETs from which we selected N = 100 models (25 from each training family) with a Pareto rank of one or less (from approximately 1200 possible choices).
The nominal, training (75 models), and prediction (25 models) errors were calculated for each objective (Table \ref{obj-table}).
Models used for prediction error calculations for a particular objective were \emph{not} trained on that objective.
The prediction likelihood was statistically significantly better for 31 of the 33 objective functions at a 95\% confidence level compared with random parameter sets generated from the nominal set (Table \ref{obj-table}).
POETs generated model families that predicted approximately 94\% of the training data with a significantly higher likelihood than a random control.
However, the specific value of any given parameter was likely not well described; the coefficient of variation (CV) for the model parameters ranged from 0.5 - 1.6, where approximately 65\% of the parameters were constrained with a CV $\leq$ 1.0 (supplementary materials Fig. \ref{fg:Para_S1}).
The most constrained parameters involved a wide-array of functions e.g., regulation of PERK, eIF2$\alpha$, ATF4, Calcineurin, BiP, CHOP and ATF6 signaling.
However, the least constrained parameters involved JNK and apoptosis interactions.
POETs identified Pareto fronts between several objectives, e.g., O13$\times$O14, O25$\times$O29, O11$\times$O29, and O27$\times$O2, in the training data  (Fig. \ref{fg:objective_plot}).
Strong Pareto fronts suggested an inability to simultaneously model different aspects of the training data.
However, fronts could also result from experimental artifacts, e.g., variation between cell-lines, time-scale differences, or from functional relationships in the data.
Globally, adaptation and alarm phase training constraints conflicted with those involving apoptosis.
For example, objectives involving caspase-7 or caspase-9 activity conflicted with phosphorylated eIF2$\alpha$ levels.
Phosphorylation of eIF2$\alpha$ by activated PERK attenuates translation, which decreases the ER folding load.
Thus, eIF2$\alpha$ phosphorylation is a key early adaptive event in UPR. On the other hand, caspase-9 is a stress-induced death marker activated only after UPR has failed to restore folding homeostasis.
Conflicts between these early and late phase markers suggested the UPR time scale was perhaps cell-line or perturbation dependent. Next, we compared model simulations with measurements of components mediating the three phases of UPR.

%For example, XBP1 mRNA measurements (O14) conflicted with CHOP protein measurements (O13), even though these data-sets were taken from the same study and were collected in the same cell-line. XBP1 splicing increased BiP levels, which in turn reduced CHOP protein levels, hence the trade-off. Lastly, in addition to fronts, we also observed strong correlation between objectives. For example, models that performed well for the CHOP protein (O11), also performed well against Procaspase-12 (O22) measurements, even though these were not in the same cell-line or from the same study. Both CHOP and Procaspase-12 are downstream of the IRE1/TRAF2/JNK signaling cascade, so these errors were directly correlated.
%\clearpage
% \subsection*{The population of UPR models recapitulated adaptation, alarm, and apoptotic events across multiple cell-lines and timescales}
%The counteracting effects are further substantiated in simulated knockout studies, wherein knockout of one ER stress transducer led to enhancement of the other branches of UPR (supplementary materials Fig. \ref{fg:Flux_Supp_Fig1}A-C).
%We compared the ability of the system to recuperate upon reversal of unfolded protein loads from the alarm phase (P4, Fig. \ref{fg:proof_of_concept_simulation} ) and the apoptosis phase (P2, Fig. \ref{fg:proof_of_concept_simulation} ). If we reduced the protein load after the cell was committed to apoptosis (as in P3), we found that the cell continued to function similar to the UPR state even upon protein load reduction after 25 hrs (Fig. \ref{fg:cross_plot_flux}C).

@@

\subsection*{Signal flow analysis suggested the UPR branches fired simultaneously}
Traditionally, it has been hypothesized that there is a sequential order for firing of the ER stress transducers in UPR.
The PERK and ATF6 branches are thought to be activated before IRE1 \cite{szegezdi2006mediators} and to largely promote ER adaptation to misfolding, while IRE1 transmits both survival and pro-apoptotic signals.
However, analysis of the UPR model ensemble suggested the three branches fired simultaneously, and that adaptation, alarm, or apoptosis was the result of counteracting effects of the three UPR signaling pathways (Fig. \ref{fg:proof_of_concept_simulation} and Fig. \ref{fg:Flux_Supp_Fig1}). UPR induction was controlled by manipulation of the generation rate of unfolded or misfolded protein (qP) in the ER compartment.
Upon induction, initially ($t\leq$ 1hr) the response was damped, marking the adaptation phase of UPR (Fig. \ref{fg:proof_of_concept_simulation}A).
Adaptation was followed by increased IRE1$\alpha$, PERK, and ATF6 activity at $t\sim$ 1 hr, marking the onset of the alarm phase.
The alarm phase was followed by a steady state at $t\sim$ 8-10 hrs, marking the onset of the apoptosis phase of UPR (Fig. \ref{fg:proof_of_concept_simulation}A).
Our analysis was substantiated further by looking at the fluxes at different phases of UPR induction (Fig. \ref{fg:proof_of_concept_simulation}B).
At P1, there was a marked increase in ATF4 and CHOP activity, ATF6 signaling and unfolded protein sensing and degradation by ERAD.
These are hallmarks of the adaptation-alarm phase of the UPR response (Fig. \ref{fg:cross_plot_flux}D).
The apoptosis phase at P2 ($t\geq$ 8-10 hrs) was marked by increased BiP regulation, enhanced ATF4 transcriptional activity, increased mitochondrial membrane permeability, and increased apoptotic fluxes.
On the other hand, if we reduced the load of unfolded protein in the adaptation-alarm phase (P4), the cell recuperated using its ERAD machinery and the regulation of BiP (Fig. \ref{fg:cross_plot_flux}E-F).
Thus, following initiation of UPR, all three sensor branches were simultaneously activated. However, it was unclear which parameters or species controlled the responses of the three branches.
We addressed this question using sensitivity and robustness analysis.

% We further investigated the effect of simulated knockout (KO) and overexpression (OX) of key proteins on the UPR system
% (supplementary materials Fig. \ref{fg:Flux_Supp_Fig1}).
% Interestingly, PERK and ATF4 KO studies revealed a slower and lower amount of BiP production ($\sim$ 50\%) as compared to WT. However, ATF6 or IRE1 KO did not affect BiP regulation as compared to WT. This highlighted the dominant role of ATF4 in regulation of BiP, which is consistent with experimental evidence \cite{ma2003delineation}. Regulation of BiP was the critical regulator of spliced XBP1 (XBP1s), which in turn acts as a key marker of progression through different stages of UPR (supplementary materials Fig. \ref{fg:Flux_Supp_Fig1}E). ATF4, cleaved ATF6, and XBP1s act as integrators of the signals coming from all the three branches of UPR and furthermore leads to regulation of BiP, thereby leading to a negative feedback or control of UPR signal. Another interesting note was the regulation of pro-apoptosis phenotype via regulation of Bcl2. PERK and ATF4 KO led to delay in the onset of apoptosis (marked by slower and lower reduction of Bcl2 levels, supplementary materials Fig. \ref{fg:Flux_Supp_Fig1}F). This effect could be attributed to the lack of CHOP mediated branch of Bcl2 regulation. On the other hand, IRE1 and CHOP KO leads to drastic reduction in apoptosis (marked by little or no change of Bcl2 levels, supplementary materials Fig. \ref{fg:Flux_Supp_Fig1}F). CHOP KO implicated the importance of CHOP in the down-regulation of Bcl2. IRE1 KO implicated the critical role of IRE1-TRAF2 mediated route of apoptosis.
% Overall signal flow analysis highlighted the extensive amount of crosstalk within the three branches of the UPR network.
% Taken together, the population of UPR models recapitulated both short- and long-timescale behavior following overload of unfolded proteins in a variety of cell types.
% It also captured the integration between multiple pathways and generated specific and testable hypothesis about the role of network components in signal propagation.
